---
title: Rows in tidyverse again
author: Jan Hlavacek
date: '2019-04-30'
slug: rows-in-tidyverse-again
categories:
  - R
  - tidyverse
tags:
  - goodness of fit
  - simulation
  - crow
---

In the [last post](https://space.lahvak.space/2019/04/14/tidyverse-rows-as-vectors/) I complained that
there is no way to apply a formula to rows of a data frame so that the formula
will consume the rows as lists or vectors.  Turns out I was wrong, and there is
a very easy way to do it!

Not only that, but I was actually really close to it, and if I was just reading
[the relevant documentation](https://www.rdocumentation.org/packages/base/versions/3.5.3/topics/dots)
bit more carefully...

As I wrote in that post, this works:

```{r eval=FALSE}
selections %>%
    mutate(chisq = pmap_dbl(., ~sum((c(..1, ..2, ..3, ..4) -
                                 expected)^2/expected))) %>%
    head()
```

I was suggesting that perhaps something like
```{r eval=FALSE}
pmap_dbl(selections, ~sum((.row - expected)^2/expected))
```
would be nice to have.  While experimenting, among other things I tries this:
```{r eval=FALSE}
pmap_dbl(selections, ~sum((... - expected)^2/expected))
```
which of course does not work, and I am not sure why I thought it would.  The
correct way, of course, is
```{r eval=FALSE}
pmap_dbl(selections, ~sum((c(...) - expected)^2/expected))
```
The `...` is, in this case, equivalent to `..1, ..2, ..3, ..4`.
Since I need `c(..1, ..2, ..3, ..4)`, I have to use `c(...)`.

## The simulation again

Anyway, let's do the whole simulation from the previous post again:

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(mosaic)
```

First enter the given information:

```{r}
labels <- fct_inorder(c("White", "Black", "Hispanic", "Other"), ordered=TRUE)

percentages <- c(72, 7, 12, 9)
names(percentages) <- labels

observed <- c(205, 26, 25, 19)
names(observed) <- labels
```

Calculate the expected frequencies and the observed $\chi^2$ score:

```{r}
expected <- 275*percentages/100

observed_chi_square <- sum((observed - expected)^2/expected)
observed_chi_square
```

Repeatedly sample, with replacement, 275 pieces of paper from a bag
representing the population, tally each sample, and record the frequencies in a
data set:

```{r}
bag <- rep(labels, percentages)
do(1000) * tally(sample(bag, 275, replace=TRUE)) -> selections
glimpse(selections)
```

Add a new column to the data set, with the $\chi^2$ scores of the simulated samples:

```{r}
selections %>%
    mutate(chisq = pmap_dbl(., ~sum((c(...) - expected)^2/expected))) ->
    selections_with_chisq
```

This is what we have now:
```{r}
head(selections_with_chisq)
```

Plot the $\chi^2$ scores of all the simulated samples, and mark the
observed $\chi^2$ on the plot with a vertical line.

```{r}
gf_histogram(~chisq, data = selections_with_chisq) %>%
    gf_vline(xintercept = observed_chi_square, color="red")
```

How many of the 1000 simulated samples had a $\chi^2$ score greater than or equal to the observed $\chi^2$ score?

```{r}
count(~(chisq >= observed_chi_square), data = selections_with_chisq)
```

## Possible simplification

I am pretty happy with this. The only thing that I think can still confuse R
beginners is the use of `.` as the first argument of `pmap_dbl`. I usually do
few examples of _tidyverse_ pipelines at the beginning of the semester, and
usually one or two of those use this somewhere, but I don't think most students
will remember those at this point.  Since we do not actually need to preserve
the original columns (the only reason I preserved them was to make the data set
look just like a worksheet they filled in while figuring out the whole idea of
$\chi^2$ scores), we can do this instead:

```{r}
selections %>% 
    pmap_df(~list(chisq = sum((c(...) - expected)^2/expected))) -> chisquares
```
and then replace `selections_with_chisq` by `chisquares` when making the
histogram and calculating the p-value.  I am not sure if this really is that
much simpler, with the `list`, and `pmap_df` not being one of the basic _dplyr_
verbs we learned at the beginning of the semester...


